---
title: "Final Project"
author: "Law Neely"
date: "2024-02-27"
output: pdf_document
---
# Load the libraries 
```{r setup, include=FALSE}
library(tidyverse)
library(dplyr)
library(MLmetrics)
library(caret)
library(ggplot2)
library(leaflet)
library(eeptools)
library(modeldata)
library(randomForest)
library(class)
```

#  partition the dataset to test with 10% of the original dataset. 
```{r, echo = FALSE}
# setwd("Users/lawrenceneely/Documents")

# read in the dataset csv file
credit_fraud <- read.csv("credit_card_fraud(v1).csv")

fraud_tibble <- as_tibble(credit_fraud)
View(credit_fraud)
```
```{r pressure, echo=FALSE}
# partition the dataset to test with 10% of the original dataset
fraud_sample <- createDataPartition(credit_fraud$is_fraud, p=0.1, list=FALSE)

strat_sample <- credit_fraud[fraud_sample, ]

#View(strat_sample)
```
# remove the unneccessary columns and create new variables such as transaction_type and category_g to convert those variables to categorical variables.  Then I added a new column for the amount range to put the amounts in different groups to make it easier to analyze. 
```{r, echo = FALSE}
# remove unncessary columns
fraud_categories <- strat_sample %>%
 select(merchant,category,amt,first,last,city,state,city_pop,job,dob,is_fraud)

# create new variable to convert is_fraud variable to categorical
fraud_categories <- fraud_categories %>% 
  mutate(transaction_type = if_else(is_fraud == 0,"legitimate", "fraud" ))

fraud_categories$transaction_type <- as.factor(fraud_categories$transaction_type)
levels(fraud_categories$transaction_type) <- c("legitimate","fraud")

# add new column to categorize the categories
categorize_categories <- function(category) {
  if (category %in% c("entertainment", "food_dining")) {
    return("Ent/Food")
  } else if (category %in% c("gas_transport", "travel")) {
    return("Transp/Trav")
  } else if (category %in% c("grocery_net", "grocery_pos")) {
    return("Groc")
  } else if (category %in% c("health_fitness", "personal_care")) {
    return("Health/PC")
  } else if (category %in% c("home", "kids_pets")) {
    return("Home/Kids_Pets")
  } else if (category %in% c("misc_net", "misc_pos")) {
    return("Misc")
  } else if (category %in% c("shopping_net", "shopping_pos")) {
    return("Shop")
  } else {
    return("Other")
  }
}

# Apply the function to create a new column
fraud_categories <- fraud_categories %>%
  mutate(category_g = sapply(category, categorize_categories))

fraud_categories <- as.data.frame(fraud_categories)


#fraud_categories$amt
breaks <- c(0, 100, 500, 1000, 5000, 10000, 20000)

fraud_categories$amt_range <- cut(fraud_categories$amt, breaks = breaks, labels = c("0-100","101-500","501-1000","1001-5000","5001-10000","10001-20000"), include.lowest = TRUE)
# logistic regression with is_fraud as outcome variable and category as predictor variable
#logreg_fraud <- glm(is_fraud ~ category, data = fraud_categories, family = binomial)

# select the most significant categories in the dataset
#significant_categories <- c("grocery_pos","misc_net","shopping_net","shopping_pos")

#significant_fraud_categories <- fraud_categories[fraud_categories$category %in% significant_categories,] 
  

#fraud_proportions <- with(significant_fraud_categories, tapply(is_fraud, category, mean))

#print(fraud_proportions)

#View(fraud_categories)

```

# bar chart to see the average fraud probability by category group. 

```{r, echo = FALSE}
# Calculate average fraud probability for each category_group
avg_prob_by_category <- fraud_categories %>%
  group_by(category_g) %>%
  summarize(avg_prob = mean(is_fraud))
```
```{r, echo = FALSE}
# Create a bar plot
ggplot(avg_prob_by_category, aes(x = category_g, y = avg_prob)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  labs(title = "Fig 1: Average Fraud Probability by Category Group",
       x = "Category Group",
       y = "Average Fraud Probability") +
  theme_minimal()

```

```{r, echo = FALSE}
# retrieve coefficients from the logistic regression
#coefficients <- coef(logreg_fraud)
#errors <- summary(logreg_fraud)$coefficients[, "Std. Error"]

#lower <- coefficients - 1.96 * errors
#upper <- coefficients + 1.96 * errors

#coef.df <- data.frame(
#  category = names(coefficients),
 # coef = coefficients,
#  lower = lower,
#  upper = upper
#)
# select top 5
#top_categories <- coef.df %>%
#  arrange(desc(abs(coef))) %>%
#  slice(1:5)

#ggplot(top_categories, aes(x = reorder(category,coef), y = coef, ymin = lower, ymax = upper)) + 
#  geom_bar(stat = "identity", fill = "skyblue") + 
#  geom_errorbar(width = 0.2, aes(ymin = lower, ymax = upper), color = "black") + 
#  coord_flip() + 
#  labs(title = "Coefficients of Shopping Fraud Categories", 
#       x = "Category", 
#       y = "Coefficient") + 
#  theme_minimal()


```
# add a new column called 'region' and split the states up into their appropriate region. I then completed a logistic regression to determine if the regions had any significance but unfortunately the regions did not.
```{r, echo = FALSE} 
fraud_categories <- fraud_categories %>%
    mutate(region = ifelse(state %in% c("AK", "CA", "OR", "WA","NV","UT","CO","WY","MT","ID","HI"), "West",
                ifelse(state %in% c("TX", "OK", "NM", "AZ"), "Southwest",
                ifelse(state %in% c("NY", "PA", "NJ", "MA","VT","NH","CT","ME","MD","DE","RI"), "Northeast",
                ifelse(state %in% c("FL", "GA", "AL", "MS","DC","WV","VA","NC","SC","LA","AR","KY","TN"), "Southeast",
                ifelse(state %in% c("IL", "OH", "MI", "WI","IA","MN","KS","NE","SD","ND","MO","IN"), "Midwest", "Other"))))))

```
```{r, echo = FALSE}

# logistic regression with is_fraud as outcome and category,region, and amt_range as predictors
logreg_fraud3 <- glm(is_fraud ~ category + region + amt_range, data = fraud_categories, family = binomial)

fraud_categories$predicted_fraud <- predict(logreg_fraud3, newdata = fraud_categories, type = "response")

#summary(logreg_fraud3)
#exp(logreg_fraud3$coefficients)
```

# create a scatterplot for the Transaction Amount Range vs Probability of Fraud to visualize the ranges with the highest probability for fraud while also adding color to symbolize the transaction category. 
```{r,echo = FALSE, warning=FALSE}
  ggplot(data = fraud_categories, aes(x = amt_range, y = predicted_fraud, color = category)) +
    geom_point() +
    geom_smooth(method = "glm", method.args = list(family = "binomial")) +
    labs(title = "Fig 2: Transaction Amount Range vs. Probability of Fraud", x = "Transaction Amount Range", y = "Probability of Fraud") +
    theme_minimal()

# Assuming 'model' is your logistic regression model
#predicted <- predict(logreg_fraud3, newdata = fraud_categories, type = "response")
#predicted_class <- ifelse(predicted > 0.5, 1, 0) # Convert probabilities to binary classes

# Calculate accuracy
#accuracy <- mean(predicted_class == fraud_categories$is_fraud)
#Accuracy(y_pred = predicted_class, y_true = fraud_categories$is_fraud)
# Create confusion matrix
#conf_matrix <- table(Actual = fraud_categories$is_fraud, Predicted = predicted_class)
#ConfusionMatrix(y_pred = predicted_class, y_true = fraud_categories$is_fraud)
#accuracy

#conf_matrix
```
# do another logistic regression to show the interaction between category and transaction amount range.

# To do that I had to first convert the amount range column to numeric. So I added a new column called amt_range_g to do that.
```{r,echo = FALSE} 
convert_range_to_numeric <- function(range) {
  # Split the range string by "-"
  range_parts <- strsplit(range, "-")[[1]]
  # Take the first part of the range as the numeric value
  numeric_value <- as.numeric(range_parts[1])
  return(numeric_value)
}
fraud_categories$amt_range <- as.character(fraud_categories$amt_range)

fraud_categories$amt_range_g <- sapply(fraud_categories$amt_range, convert_range_to_numeric)

#View(fraud_categories)

```
# Then I completed another logistic regression instead using category_group * amt_range_g to see the interactions of the two variables. 

```{r, echo=FALSE,warning=FALSE}
# logistic regression with is_fraud as outcome but category and amount range groups as the predictor variables.
logreg_fraud_grouped <- glm(is_fraud ~ category_g * amt_range_g, data = fraud_categories, family = binomial)
```

# From that logistic regression I created a plot to show the coefficients from the logistic regression.
# The coefficients do show that the interaction between category and transaction amount range can actually affect fraud probability especially with Shopping and Miscellaneous categories showing higher probabilities of fraud.

```{r, echo = FALSE}
# Extract coefficients
coefficients <- summary(logreg_fraud_grouped)$coefficients

# Create a dataframe from coefficients
coefficients_df <- as.data.frame(coefficients)

# Plot coefficients
ggplot(coefficients_df, aes(x = row.names(coefficients_df), y = Estimate)) +
  geom_point() +
  geom_errorbar(aes(ymin = Estimate - 1.96 * `Std. Error`, ymax = Estimate + 1.96 * `Std. Error`), width = 0.2) +
  coord_flip() +
  labs(title = "Fig 3: Coeffs from Logistic Regression Model",
       x = "Variables",
       y = "Estimate")
```


